Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-06-29 fully-abstract-comp/popl2019/paper.tex
First sketch of POPL submission.

\documentclass[sigplan,screen]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage{cleveref}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage[utf8]{inputenc}

% Copyright
\setcopyright{rightsretained}
\acmDOI{}
\acmISBN{}
\acmConference[PriSC]{Workshop on Principles of Secure Compilation}{2018}{Los Angeles} 
\acmYear{2018}
\copyrightyear{2018}
\settopmatter{printacmref=false}

% \acmPrice{15.00}

%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

\begin{document}
\title{Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities}
% \titlenote{Produces the permission block, and
%   copyright information}
\subtitle{Extended Abstract}
% \subtitlenote{The full version of the author's guide is available as
%   \texttt{acDat zou in dmart.pdf} document}

\author{Lau Skorstengaard}
% \orcid{}
\affiliation{%
  \institution{Aarhus University}
}
\email{lau@cs.au.dk}

\author{Dominique Devriese}
\orcid{0000-0002-3862-6856}
\affiliation{%
  \institution{imec-DistriNet, KU Leuven}
}
\email{dominique.devriese@cs.kuleuven.be}

\author{Lars Birkedal}
\affiliation{%
  \institution{Aarhus University}
}
\email{birkedal@cs.au.dk}

% The default list of authors is too long for headers}
% \renewcommand{\shortauthors}{B. Trovato et al.}

\begin{abstract}
We propose and study a new calling convention that provably enforces well-bracketed control flow and local state encapsulation on a capability machine.
The calling convention is based on linear capabilities, a type of capabilities that are prevented by the hardware from being duplicated.
In addition to designing and formalising this new calling convention, we also contribute a new way to formalise and prove that it effectively enforces well-bracketed control flow and local state encapsulation, building on the concept of fully abstract compilation.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>10010520.10010553.10010562</concept_id>
%   <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010575.10010755</concept_id>
%   <concept_desc>Computer systems organization~Redundancy</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10010520.10010553.10010554</concept_id>
%   <concept_desc>Computer systems organization~Robotics</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>10003033.10003083.10003095</concept_id>
%   <concept_desc>Networks~Network reliability</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>  
% \end{CCSXML}

% \ccsdesc[500]{Computer systems organization~Embedded systems}
% \ccsdesc[300]{Computer systems organization~Redundancy}
% \ccsdesc{Computer systems organization~Robotics}
% \ccsdesc[100]{Networks~Network reliability}

% \keywords{ACM proceedings, \LaTeX, text tagging}

% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{This is a teaser}
%   \label{fig:teaser}
% \end{teaserfigure}

@@

\section{Linear Capabilities, not Local}
What we essentially want to achieve in a secure calling convention is to give an invoked function access to a stack pointer and a return pointer, but only temporarily: during the execution of the invoked function. 
In our original approach, we make both pointers local, so that the CHERI processor guarantees, essentially, that they can only be stored in registers or on the stack.
To prevent an invoked function from transferring the pointers across outside the lifetime of its stack frame, it then suffices to block only these two communication channels, rather than the entire memory.
In other words, we need to clear registers and the stack before passing control to untrusted code, to prevent it from obtaining access to stack and return pointers for other stack frames than its own.
The requirement for clearing registers is unproblematic, but clearing the entire stack upon every boundary crossing is a non-trivial requirement.
We speculate that it can be made efficient with custom hardware extensions\footnote{Specifically, we imagine a kind of pre-L0 processor cache that does not retain values but only remembers that a certain range has been zeroed out.}, but this is not obvious.

An alternative that does not have this requirement, relies on \emph{linear capabilities}, rather than local ones.
Although details in the literature are scarce, linear capabilities have already been implemented in the SAFE machine~\citep{amorim_verified_2014}\footnote{Private communication with Catalin Hrit\c{c}u.} and considered for implementation in CHERI\footnote{Private communication with Robert Watson.}.
The idea is that all capabilities are marked as either regular or linear.
Linear ones are subject to special hardware-enforced restrictions, that prevent them from being duplicated.
For example, when linear capabilities are moved to a new location in registers or memory, they are erased from their old location.
A few special instructions are needed to make this workable, e.g.\ a split instruction that splits a memory capability for a region of memory into disjoint capabilities for two parts of the region and a splice instruction that does the reverse.
This linearity is a powerful tool: when passing a linear capability to untrusted code and receiving the same capability back, we are sure that they cannot have stored a copy of it.
Similarly, when we receive, from untrusted code, a linear capability that addresses memory that we know is only ever addressed by linear capabilities, then we are sure that the untrusted code no longer has access that memory.

As such, linear capabilities can be used as an alternative to local capabilities to enforce well-bracketed control flow.
We cannot provide full details here for space reasons, but the idea is essentially the following.
By passing stack and return pointers as linear capabilities, and requiring that invoked functions hand in their stack pointer upon return, untrusted code is prevented from holding on to stack and return pointers past their intended lifetime, essentially because they can only ever have one copy and they have to hand it in when this lifetime ends.

\section{Formalising well-bracketed control flow}
The second improvement we are making is related to the properties of well-bracketed control flow and local state encapsulation.
To establish these properties, our previous work followed previous work \citep{dreyer_impact_2010} in providing a sound reasoning technique and using it to prove soundness of challenging example programs that rely on the properties.
In the current work, we are investigating an alternative way to formalise the two properties, based on the notion of fully abstract compilation \citep{abadi_protection_1999}.

We intend to start by formalising the assembly language for a capability machine with linear capabilities (the target language).
Then, we will introduce the source language: a variant of this target language that features a primitive stack, built into the operational semantics, as well as additional instructions like call.
While the target language is a standard assembly language without even the notion of a stack, the source language does have a stack, and control flow well-bracketedness and local state encapsulation follow more directly from the operational semantics.
We will formalise our calling convention in the definition of a compiler from the source to the target language.
Additionally, we conjecture that this compiler can be proved fully abstract, thus formally expressing and establishing security of the calling convention.

Interestingly, the proof of full abstraction for this compiler is expected to use a very simple back-translation, that simply maps every target language instruction to the corresponding instruction in the source.
This back-translation works because we construct our source language in such a way that it has direct counterparts to every target language value.
Specifically, the source language contains a representation of stack and return pointers as unspecified abstract values whose behavior corresponds to their target counterparts, but in a way that guarantees the intended properties of well-bracketed control flow and local state encapsulation.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
% \acmJournal{PACMPL}
% \acmVolume{1}
% \acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
% \acmArticle{1}
% \acmYear{2018}
% \acmMonth{1}
% \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
% \startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

% %% Some recommended packages.
% \usepackage{booktabs}   %% For formal tables:
%                         %% http://ctan.org/pkg/booktabs
% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%                         %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title{Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities}
% \titlenote{with title note}
% \subtitle{Fully abstract overlay semantics}
% \subtitlenote{with subtitle note}

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

\author{Lau Skorstengaard}
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \institution{Aarhus University}
  \country{Denmark}                    %% \country is recommended
}
\email{lask@cs.au.dk}

%% Author with two affiliations and emails.
\author{Dominique Devriese}
\orcid{0000-0002-3862-6856}             %% \orcid is optional
\affiliation{
  \institution{KU~Leuven}           %% \institution is required
  \country{Belgium}                   %% \country is recommended
}
\email{dominique.devriese@cs.kuleuven.be}         %% \email is recommended

\author{Lars Birkedal}
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \institution{Aarhus University}
  \country{Denmark}                    %% \country is recommended
}
\email{birkedal@cs.au.dk}

\begin{abstract}
We propose and study a new calling convention that provably enforces well-bracketed control flow and local state encapsulation on a capability machine.
The calling convention is based on linear capabilities, a type of capabilities that are prevented by the hardware from being duplicated.
In addition to designing and formalising this new calling convention, we also contribute a new way to formalise and prove that it effectively enforces well-bracketed control flow and local state encapsulation, using what we call a fully abstract overlay semantics.
\end{abstract}

% %% 2012 ACM Computing Classification System (CSS) concepts
% %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10011007.10011006.10011008</concept_id>
% <concept_desc>Software and its engineering~General programming languages</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003521.10003525</concept_id>
% <concept_desc>Social and professional topics~History of programming languages</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

\keywords{fully abstract compilation, secure compilation, capability machines, linear capabilities, well-bracketed control flow, stack frame encapsulation, overlay semantics}

@@

% %% Acknowledgments
% \begin{acks}                            %% acks environment is optional
%                                         %% contents suppressed with 'anonymous'
%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%   %% acknowledge financial support and will be used by metadata
%   %% extraction tools.
%   This material is based upon work supported by the
%   \grantsponsor{GS100000001}{National Science
%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%   conclusions or recommendations expressed in this material are those
%   of the author and do not necessarily reflect the views of the
%   National Science Foundation.
% \end{acks}

\section{Linear Stack and Return Capabilities}
\begin{itemize}
\item informally explain how an adversary may try to abuse stack and return pointers
\item informally explain how we prevent this using linear capabilities
\item use the tikz pictures from the PriSC presentation to explain all of this
\end{itemize}
\section{A Capability Machine with Sealing and Linear Capabilities}
\begin{itemize}
\item present our target language and its operational semantics (excerpts)
\item mention roughly what components look like
\end{itemize}
\section{Formulating Security with a Fully Abstract Overlay Semantics}
\begin{itemize}
\item present our source language, its operational semantics (excerpts)
\item mention our assumption of reasonability
\item present the full abstraction theorem.
\item sketch high-level structure of the proof
\end{itemize}
\section{Discussion}
\begin{itemize}
\item mention that tail calls are supported through xcall
\item explain how fully abstract overlay semantics could form one pass of a verified secure compiler.
\end{itemize}
\section{Related Work}
\begin{itemize}
\item CFI and friends
\item our ESOP 2018 paper \citep{skorstengaard_reasoning_2017}
\item undefined behavior and full abstraction \citep{juglaret_beyond_2016}
\end{itemize}
\bibliography{references}

%% Appendix
% \appendix
% \section{Appendix}

% Text of appendix \ldots
